CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-06-30 script/en/chapters/cryptomethods.tex
be updates

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product-algorithm\index{Cascade cipher}%
\footnote{%
A cipher text can be used as input for another encryption algorithm. 
A cascade cipher is build up as a composition of different encryption 
transformations.\\
Cascading does not always improve the security.\\
This process is also used within modern algorithms:
They usually combine simple and, considered at its own, cryptologically
relatively unsecure single steps in several rounds into an efficient 
overall procedure.  Most block ciphers (e.g. DES, IDEA) are cascade ciphers.\\
Also serial usage of the same cipher with different keys (like with Triple-DES)
is called cascade cipher. 
}
does not exponentially increase with the number of rounds.

>>>>

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product-algorithm\index{Cascade cipher}%
\footnote{%
A cipher text can be used as input for another encryption algorithm. 
A cascade cipher is build up as a composition of different encryption 
transformations.\\
Cascading does not always improve the security.\\
This process is also used {\em within} modern algorithms:
They usually combine simple and, considered at its own, cryptologically
relatively unsecure single steps in several rounds into an efficient 
overall procedure.  Most block ciphers (e.g. DES, IDEA) are cascade ciphers.\\
Also serial usage of the same cipher with different keys (like with Triple-DES)
is called cascade cipher. 
}
does not exponentially increase with the number of rounds.

### 2003-06-30 script/en/chapters/digitalsignatures.tex
be updates

% --------------------------------------------------------------------------
\subsection{Hash functions}
\index{Hash function}
A cryptographic \index{Hash function} {\em hash function}\footnote{%
Secure hash algorithms compute a condensed representation of 
electronic data (message).
When a message is input to a hash algorithm, the result is an output called
a message digest. The message digests typically range in length 
from 128 to 512 bits, depending on the algorithm. 
Secure hash algorithms are typically used with
other cryptographic algorithms, such as digital signature algorithms and
keyed-hash message authentication codes, or in the generation of random
numbers (bits).
}
maps a message of any
length to a string of characters with a constant size,
the \index{Hash value} hash value. It should be practically impossible, for a
given number, to find a message that has precisely this number as hash value.
Furthermore, it should be practically impossible to find two messages with the
same hash value\index{collisions}. 
In both cases the final signature procedure would have weaknesses.

@@

The output length of the SHA algorithms was enhanced because of the
possibility of birthday attacks\index{Birthday attack}\index{collisions}:
these make n-bit AES and a 2n-bit hash roughly equivalent: \\
- 128-bit AES -- SHA-256 \\
- 192-bit AES -- SHA-384 \\
- 256-bit AES -- SHA-512.

With CrypTool\index{CrypTool} v1.3.04 you can comprehend the birthday attack
\index{Birthday attack} on digital signatures: \\
using menu {\bf Indiv. Procedures \textbackslash{} 
Attack on the hashvalue of the digital signature}.

@@

Another name for this type of attack is \index{Man-in-the-middle attack} ``man-­in-­the-­middle attack''. Users are promised
protection against this type of attack by public­key
certification, which is intended to guarantee the
\index{Authenticity} authenticity of public keys. The most common
certification method is the X.509 standard.

@@

\bibitem[PKCS1]{PKCS1} RSA Laboratories\\ 
    \index{PKCS\#1} \index{RSA Laboratories}
    {\em PKCS \#1 v2.1 Draft 3: RSA Cryptography Standard}, April 19, 2002.

\bibitem[X.509v3]{X.509v3} ITU-T \index{X.509} \index{ITU-T} \index{ISO/IEC 9594-8}
    {\em X.509 (1993) Amendment 1: Certificate Extensions, The Directory
    Authentication Framework},\\ 
    International Telecommunication Union, Geneva, Switzerland, July 1995\\
    (equivalent to Amendment 1 to ISO/IEC 9594-8).

\bibitem[FIPS180-2]{FIPS180-2} U.S. Department of Commerce/N.I.S.T. 
    \index{FIPS180-2} \\
    {\em Secure Hash Standard (SHS)}, August 1, 2002.\\
    (FIPS 180-2 will supersede FIPS 180-1 beginning February 1, 2003.)

\bibitem[FIPS186]{FIPS186} U.S. Department of Commerce/N.I.S.T.
    \index{FIPS186} \\
    {\em Entity authentication using public key cryptography}, February 18, 1997.\\
    No more valid.

\bibitem[FIPS186-2]{FIPS186-2} U.S. Department of Commerce/N.I.S.T. 
    \index{FIPS186-2} \\
    {\em Digital Signature Standard (DSS)}, January 27, 2000.\\
    Change Note: October 5, 2001.\\
    http://csrc.nist.gov/publications/fips/fips186-2/fips186-2-change1.pdf
\bibitem[X.509]{X.509}
    \index{X.509}
    {\em ITU-T Recommendation X.509 (1997 E): Information Technology -- 
    Open Systems Interconnection -- The Directory: Authentication Framework}, June 1997.

>>>>

% --------------------------------------------------------------------------
\subsection{Hash functions}
\index{Hash function}
A cryptographic \index{Hash function} {\em hash function}\footnote{%
Secure hash algorithms compute a condensed representation of 
electronic data (message).
When a message is input to a hash algorithm, the result is an output called
a message digest. The message digests typically range in length 
from 128 to 512 bits, depending on the algorithm. 
Secure hash algorithms are typically used with
other cryptographic algorithms, such as digital signature algorithms and
keyed-hash message authentication codes, or in the generation of random
numbers (bits).
}
maps a message of any
length to a string of characters with a constant size,
the \index{Hash value} hash value. It should be practically impossible, for a
given number, to find a message that has precisely this number as hash value.
Furthermore, it should be practically impossible to find two messages with the
same hash value\index{Collision}. 
In both cases the final signature procedure would have weaknesses.

@@

The output length of the SHA algorithms was enhanced because of the
possibility of birthday attacks\index{Attack!birthday}\index{Collision}:
% \index{Birthday attack}
these make n-bit AES and a 2n-bit hash roughly equivalent: \\
- 128-bit AES -- SHA-256 \\
- 192-bit AES -- SHA-384 \\
- 256-bit AES -- SHA-512.

With CrypTool\index{CrypTool} v1.3.04 you can comprehend the birthday attack
\index{Attack!birthday} on digital signatures: \\
using menu {\bf Indiv. Procedures \textbackslash{} 
Attack on the hashvalue of the digital signature}.

@@

Another name for this type of attack is \index{Attack!man-in-the-middle}
% \index{Man-in-the-middle attack}
``man-­in-­the-­middle attack''. Users are promised
protection against this type of attack by public­key
certification, which is intended to guarantee the
\index{Authenticity} authenticity of public keys. The most common
certification method is the X.509 standard.

@@

\bibitem[PKCS1]{PKCS1} RSA Laboratories,\\ 
    \index{PKCS\#1} \index{RSA Laboratories}
    {\em PKCS \#1 v2.1 Draft 3: RSA Cryptography Standard}, \\
    April 19, 2002.

\bibitem[X.509v3]{X.509v3} ITU-T, \\
    \index{X.509} \index{ITU-T} \index{ISO/IEC 9594-8}
    {\em X.509 (1993) Amendment 1: Certificate Extensions, The Directory
    Authentication Framework},\\ 
    International Telecommunication Union, Geneva, Switzerland, July 1995\\
    (equivalent to Amendment 1 to ISO/IEC 9594-8).

\bibitem[X.509]{X.509} ITU-T, \\
    \index{X.509}
    {\em ITU-T Recommendation X.509 (1997 E): Information Technology -- 
    Open Systems Interconnection -- The Directory: Authentication Framework},\\
    June 1997.

\bibitem[FIPS180-2]{FIPS180-2} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS180-2} \\
    {\em Secure Hash Standard (SHS)}, \\
    August 1, 2002.\\
    (FIPS 180-2 will supersede FIPS 180-1 beginning February 1, 2003.)

\bibitem[FIPS186]{FIPS186} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS186} \\
    {\em Entity authentication using public key cryptography}, \\
    February 18, 1997.\\
    No more valid.

\bibitem[FIPS186-2]{FIPS186-2} U.S. Department of Commerce/N.I.S.T. ,
    \index{FIPS186-2} \\
    {\em Digital Signature Standard (DSS)}, \\
    January 27, 2000. Change Note: October 5, 2001.\\
    http://csrc.nist.gov/publications/fips/fips186-2/fips186-2-change1.pdf

### 2003-06-30 script/en/chapters/ellipticcurves.tex
be updates

Using the DSA signature scheme, one can proceed as follows: The signing party chooses a (non-trivial) number $s\in{\mathbb Z}_q$, which will be the private key, and publishes $q$, $G$ and $R=sG$. We note that $s$ cannot be obtained from $G$ and $R$ are not sufficient --- a fact on which the security of the signature scheme is based.

@@

The procedure implements the cryptographic primitives for generating and
verifying signatures for the variations of Nyberg-Rueppel signatures and
\index{DSA} DAS signatures based on elliptic curves (in accordance with draft
no. 8 of the IEEE P1363 work group). This was done in collaboration with the
Secude GmbH --- using the above library and the Secude SDK.
\index{Secude}

@@

        \bibitem[Cassels1991]{Cassels1991} J. W. S. Cassels, {\em Lectures on elliptic curves}, Cambridge University Press, 1991, 143 pages.
	\bibitem[Koblitz1984]{Koblitz1984}N. Koblitz, {\em Introduction to elliptic curves and modular forms}, Graduate Texts in Mathemathics, Springer-Verlag, 1984.
	\bibitem[Koblitz1998]{Koblitz1998}N. Koblitz, 
	{\em Algebraic aspects of Cryptography. With an appendix on 
	Hyperelleptic curves by Alfred J. Menezes, Yi Hong Wu and Robert 
	J. Zuccherato}, Springer-Verlag, 1998, 206 pages.
	\bibitem[Menezes1993]{Menezes1993}A. J, Menezes, {\em Elliptic curve public key cryptosystems}, Kluwer Academic Publishers, 1993.
        \bibitem[Lenstra1987]{Lenstra1987} H.W. Lenstra \index{Lenstra 1987} \\
                {\em Factoring integers with elliptic curves}, Annals of Mathematics 126, pp. 649-673, 1987.
	\bibitem[Lenstra1999]{Lenstra1999} Arjen K. Lenstra, Eric R. Verheul  \index{Lenstra/Verheul 1999} \\
         	{\em Selecting Cryptographic Key Sizes (1999)}, Journal of Cryptology: the journal of the International Association for Cryptologic Research \\
		\href{http://www.cryptosavvy.com/cryptosizes.pdf}{\texttt{http://www.cryptosavvy.com/cryptosizes.pdf}}
	\bibitem[Silverman1986]{Silverman1986} J. Silverman \\ {\em The Arithmetic of Elliptic Curves}, Springer-Verlag, 1986. \index{Silverman 1986}
	\bibitem[Silverman1992]{Silverman1992}J. Silverman, {\em The arithmetic of elliptc curves}, Graduate Texts in Mathemathics, Springer-Verlag, 1992.
	\bibitem[SivermanTate1992]{SivermanTate1992}J. Silverman, J. Tate, {\em Rational points on elliptic curves}, Springer-Verlag, 1992.

>>>>

Using the \index{DSA} DSA signature scheme, one can proceed as follows: The signing party chooses a (non-trivial) number $s\in{\mathbb Z}_q$, which will be the private key, and publishes $q$, $G$ and $R=sG$. We note that $s$ cannot be obtained from $G$ and $R$ are not sufficient --- a fact on which the security of the signature scheme is based.

@@

The procedure implements the cryptographic primitives for generating and
verifying signatures for the variations of Nyberg-Rueppel signatures and
\index{DSA} DSA signatures based on elliptic curves (in accordance with draft
no. 8 of the IEEE P1363 work group). This was done in collaboration with the
Secude GmbH --- using the above library and the Secude SDK.
\index{Secude}

@@

    \bibitem[Cassels1991]{Cassels1991} J. W. S. Cassels,
	\index{Cassels 1991} \\
        {\em Lectures on elliptic curves},\\
	Cambridge University Press, 1991, 143 pages.
    \bibitem[Koblitz1984]{Koblitz1984}N. Koblitz, 
	\index{Koblitz 1984} \\
        {\em Introduction to elliptic curves and modular forms},\\
	Graduate Texts in Mathemathics, Springer-Verlag, 1984.
    \bibitem[Koblitz1998]{Koblitz1998}N. Koblitz,  
	\index{Koblitz 1998}\\
	{\em Algebraic aspects of Cryptography. With an appendix on 
	Hyperelleptic curves by Alfred J. Menezes, Yi Hong Wu and Robert 
	J. Zuccherato}, \\
	Springer-Verlag, 1998, 206 pages.
    \bibitem[Menezes1993]{Menezes1993}A. J., Menezes, 
	\index{Menezes 1993} \\
        {\em Elliptic curve public key cryptosystems},\\
	Kluwer Academic Publishers, 1993.
    \bibitem[Lenstra1987]{Lenstra1987} H.W. Lenstra, 
        \index{Lenstra 1987} \\
        {\em Factoring integers with elliptic curves}, \\
	Annals of Mathematics 126, pp. 649-673, 1987.
    \bibitem[Lenstra1999]{Lenstra1999} Arjen K. Lenstra, Eric R. Verheul
        \index{Lenstra/Verheul 1999} \\
        {\em Selecting Cryptographic Key Sizes (1999)},\\
	Journal of Cryptology: the journal of the International 
	Association for Cryptologic Research \\
	\href{http://www.cryptosavvy.com/cryptosizes.pdf}
	{\texttt{http://www.cryptosavvy.com/cryptosizes.pdf}}
    \bibitem[Silverman1986]{Silverman1986} J. Silverman, 
        \index{Silverman 1986}\\
        {\em The Arithmetic of Elliptic Curves},\\
	Springer-Verlag, 1986.
    \bibitem[Silverman1992]{Silverman1992}J. Silverman,
	\index{Silverman 1992} \\
        {\em The arithmetic of elliptc curves},\\
	Graduate Texts in Mathemathics, Springer-Verlag, 1992.
    \bibitem[SivermanTate1992]{SivermanTate1992}J. Silverman, J. Tate,
	\index{Siverman/Tate 1992} \\
        {\em Rational points on elliptic curves},\\
	Springer-Verlag, 1992.

### 2003-06-30 script/en/chapters/numbertheory.tex
be updates

\newpage
%\hypertarget{Kap_3}
\hypertarget{Kapitel_3}{}

\section{Introduction to Elementary Number Theory with Examples}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003) \\
\label{Kapitel_3}

@@

For more details on prime numbers (e.g. how ``Fermat's Little Theorem'' can
be used to test extremely large numbers to determine whether they are prime),
please refer to the article on \hyperlink{Kapitel_2}{\ref{Label_Kapitel_2}, 
Prime numbers} in this script.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Groups and modular arithmetic in $\mathbb{Z}_n$ and $ \mathbb{Z}_n^*$}
\index{Group}
Mathematical ``{\em groups}'' play a decisive role in number theory and cryptography. We only talk of groups if, for a 
defined set and a defined relation (an operation such as addition or multiplication), the following properties are fulfilled:

@@

The task of the researchers from Bonn was not initiated by a challenge, but
they wanted to find the last prime factors of the integer $2^{953} - 1$
(see ``Wanted List'' of the Cunningham Project\index{Cunningham project}%
\footnote{\href{http://www.cerias.purdue.edu/homes/ssw/cun/}{\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}}).

@@

\bibitem[GISA2002]{GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency) \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}
    {\tt http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf} \\
    A statement on these recommendations: \\
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{http://www.secorvo.de/publikat/}}\\
    \vspace*{1cm}\href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{stellungnahme-algorithmenempfehlung-020307.pdf}}

@@

I would like to take this opportunity to thank 
\begin{itemize}
  \item {Mr.\ Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading this article and
         for helping with TeX. \\
	 Mr.\ Koy designed and developed the functions and the complex
	 dialog box of the RSA cryptosystem, which enable you to execute
	 the RSA samples of this article.}
  \item {J\"org Cornelius Schneider for his TeX support und for the many 
         cases where he helped when facing programming or design problems.}
  \item {Dr. Illies for pointing me to Pari-GP \index{Pari-GP}.}
\end{itemize}

>>>>

\newpage
\hypertarget{Kapitel_3}{}

\section{Introduction to Elementary Number Theory with Examples}
\label{Kapitel_3}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003) \\

@@

For more details on prime numbers (e.g. how ``Fermat's Little Theorem'' can
be used to test extremely large numbers to determine whether they are prime),
please refer to the article on \hyperlink{Kapitel_2}{prime numbers, 
\ref{Label_Kapitel_2}} in this script.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Groups and modular arithmetic in $\mathbb{Z}_n$ and $\mathbb{Z}_n^*$}
\index{Group}
Mathematical ``{\em groups}'' play a decisive role in number theory and cryptography. We only talk of groups if, for a 
defined set and a defined relation (an operation such as addition or multiplication), the following properties are fulfilled:

@@

The task of the researchers from Bonn was not initiated by a challenge, but
they wanted to find the last prime factors of the integer $2^{953} - 1$
(see ``Wanted List'' of the Cunningham Project
\index{Cunningham project}\footnote{%
Cunningham project: \href{http://www.cerias.purdue.edu/homes/ssw/cun/}
                 {\texttt{http://www.cerias.purdue.edu/homes/ssw/cun/}}}).

@@

\bibitem[GISA2002]{GISA2002} \index{GISA 2002}
    GISA (German Information Security Agency), \\
    {\em Recommendation for key length selection}, Bonn, Sep. 2002, \\
    \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf}
    {\tt http://www.bsi.bund.de/esig/basics/techbas/krypto/bund02v7.pdf} \\
    A statement on these recommendations: \\
    \href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{http://www.secorvo.de/publikat/}}\\
    \vspace*{1cm}\href{http://www.secorvo.de/publikat/stellungnahme-algorithmenempfehlung-020307.pdf}
    {\texttt{stellungnahme-algorithmenempfehlung-020307.pdf}}

@@

I would like to take this opportunity to thank 
\begin{itemize}

  \item {Mr.\ Henrik Koy for making many very useful suggestions, 
         for the very constructive proof-reading this article and
         for helping with TeX. \\
	 Mr.\ Koy designed and developed in his leisure time the 
	 functions and the complex dialog box of the RSA cryptosystem, 
	 which enable you to execute the RSA samples of this article.}

  \item {J\"org Cornelius Schneider for his TeX support und for the many 
         cases where he helped when facing programming or design problems.}

  \item {Dr. Illies for pointing me to Pari-GP \index{Pari-GP}.}
